---
title: "Visualizing and Describing Networks"
---

## Networks with igraph 

First, make sure you have loaded package *igraph*. 
```{r}
library(igraph)
```

First, create a simple adjacency matrix with three rows and three columns 
```{r}
mat1 <- matrix(c(0, 1, 0, 0, 0, 1, 1,0, 0), nrow=3, ncol=3) ### matrix function 
mat1
```


Use igraph function graph_from_adjacency_matrix() to create a network object from your graph. Use plot function to plot. 
```{r}
mat2 <- graph_from_adjacency_matrix(mat1)
plot(mat2, edge.arrow.size = 1) ## set the size of the arrows 
```

Alternatively, create the network by telling igraph what links you have 

```{r}
mat3 <- graph( edges=c(1,3, 3,2, 2,1), n=3, directed=T ) # use graph function and list edges 
plot(mat3, edge.arrow.size = 1)
```

## Network Aestetics
Many parts of a network can be sized and colored to help communicate results more clearly. 

```{r}
plot(mat3, edge.arrow.size = 1, vertex.color = "purple", vertex.size = 20, vertex.label.cex = 2, vertex.label.dist = 3.5)

```





### Describing networks


Use igraph "graph" function to plot directly as igraph object. 
```{r}
g1 <- graph( edges=c(1,3, 3,2, 2,1), n=3, directed=T ) 
g1
plot(g1, edge.arrow.size = .5)
```

Undirected
```{r}
g2 <- graph(edges = c(1,3, 1,2, 1,4, 2,4), n = 4, directed = F)
g2
plot(g2, edge.arrow.size = .5)
```

Graph Density, diameter, mean path length
```{r}
igraph::graph.density(g2) #graph density 
diameter(g2) # diameter
mean_distance(g2) ##mean path length
```

directed graph
```{r}
g3 <- graph(edges = c(1,3, 1,2, 1,4, 2,4), n = 4, directed = T)
g3
plot(g3, edge.arrow.size = .7)
```


Toy example for betweenness centrality 
```{r}
g5 <- graph(edges = c(1,2, 2,3, 3,4, 4,5), n = 5, directed = F)
plot(g5, vertex.label.cex = .0001)
```


```{r}
igraph::closeness(g2, vids = V(g2), normalized = TRUE)
igraph::closeness(g2, vids = V(g2), normalized = FALSE)
igraph::betweenness(g2)
```

Geodesic
```{r}
Geodesic <- igraph::distances(g2)
Geodesic ## geodesic matrix
hist(Geodesic)
```

degree distribution
```{r}
Degree_Distribution <- igraph::degree(g3, mode = "total")
hist(Degree_Distribution, xlab = "Node Degree") #all

degIn <- igraph::degree(g3, mode = "in") #in
degOut <- igraph::degree(g3, mode = "out") #out

hist(Degree_Distribution, xlab = "Node Degree") 
hist(degIn, xlab = "Node Degree")
hist(degOut, xlab = "Node Degree")
```


```{r}
igraph::vertex_connectivity(g2)
igraph::transitivity(g2)
```


back to our class example 
```{r}
Edges <- data.frame(
  From =c("Jim", "Jim", "Jim", "Jim", "Carole", "Carole", "Carole", "Joe", "Joe", "Joe", "Michelle", "Michelle", "Jen", "Jim"),
  To = c("Carole", "Joe", "Mark", "Jill", "Jen", "Pete", "Paul", "Jill", "Cam", "Kate", "Jill", "Tim" , "Cam", "Jess"))

Nodes <- data.frame(
  Name = c("Jim", "Carole", "Joe", "Michelle", "Jen", "Pete", "Paul", "Tim", "Jess", "Mark", "Jill", "Cam",                               "Kate"),
  Experience = c(8.5, 6.5, 4.0, 1.0, 3.0, 10.0, 5.0, 5.0, 5.0, 1.0, 1.0, 6.0, 6.0))


AsMatrix <- graph_from_data_frame(d = Edges, vertices = Nodes, directed = FALSE)
AsMatrix


plot(AsMatrix)


Degree_Distribution <- igraph::degree(AsMatrix, mode = "total")
```

same number of nodes and links
```{r}
new1 <- sample_gnm(13, 14, directed = FALSE, loops = FALSE)
h1<- igraph::degree(new1)
hist(h1)
```


run that n number of times 
```{r}
degamat <- NULL
n <- 50
for(i in 1:n){
  newmatrix <- sample_gnm(13,14, directed = FALSE, loops = FALSE)
  degmat <- igraph::degree(newmatrix)
  degamat<-rbind(degamat,degmat)
}
degamat

hist(degamat, xlim = c(0,7), breaks = 7)
```

```{r}
par(mfrow=c(1,1),
    mar=c(2,2,2,2))
hist(Degree_Distribution, xlab = "Node Degree", xlim = c(0,7), breaks = 3, main = "Observed")
hist(degamat, xlim = c(0,7), breaks = 7, xlab = "Node Degree", main = "Simulated")
```
